---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

# Introduction

The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, harbour constructions and, as is the focus of this study, Stone Age sites.

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). The method of shoreline dating has been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Post-glacial isostatic rebound leading to shoreline fall

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated various and multiple sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @crema2010; @crema2012; @lewis2021], a similar approach is adopted here.

# Background

Relative sea-level is so named because the parameters of interest is the elevation of the surface of the sea relative to land. Variation in this relative distance follows from effects that can be divided into three main categories. The first of these is changes to the eustatic sea-level, which is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

sea level is the distance from the solid surface of the Earth to the mean sea level, as averaged over time, and is also known as the geoid.

Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory---a process that also continues to this day. In other areas of Norway the situation is sometimes more complicated. On the outer coast, the relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed. In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle. Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span. As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.

The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this. In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was *not* shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the site which would then have been accessible by means of watercraft. This line of reasoning is often also used in survey projects, which typically do not produce material suitable for radiocarbon dating. The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and collegues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites. However, in several of the cases where there was a discrepancy, the

Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displaceent curves and radiocarbon dates, it is appropriate to evaulate this relationship by means of statistical equivalence tests. This effectively shifts the common direction of the null-hypothesis, giving H0: The date given by means of radiocarbon dating and by shoreline displacement are different. Ha: The date given by means of radiocarbon dating and by shoreline displacement are the same. While this consitutes the formal dimension of the study, there must necessarily also be a qualitative and more subjectivly driven dimension as well. As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates. Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report. While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.

This study necessaitated the compilation of multiple relative sea-level curves developed for different parts of the region. Most of the curves come provided with a confidence interval which is largely based on an evaluation by the geologists who developed the curve. The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples, as well as direction and adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few. The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods. However, any reinterpretation and standardisation of the curves is also hampered by a range of issues. The original data or necessary details about the sampling processes are not always available, and a reinterpretation based soley on radiometric dates, without taking into account the considerations and evaluations of the field geologists can be entirely misleading. In a rudimentary attempt to account for these differences when compiling the curves, the width of the confidence band for the sea level at a single date across the entire region is given as the largerst uncertainty of any one of the curves at that point in time. This averages to an uncertainty of +/- calendar years. This uncertainty is uniformly distributed within the confidence band.

# Data

All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive, including excavated sites without radiocarbon dates, and radiocarbon dates to periods other than the Stone Age.

The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west. These number a total nrow() sites, of which nrow() have radiocarbon dates. Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below). Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the Museum of Cultural History. Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.

# Methods

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shore-lines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena.

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the meter distance between the isobases. This was done for both ends of the confidence intervals. To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve. The result of this process is shown for an example site in Figure.

For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve using OxCal v. . All dates not younger than 1700 BCE with 99.7 % uncertainty were then excluded from further analysis. Radiocarbon dates associated with each site were then grouped if they overlaped with 99.7 % uncertainty, effectively taken to represent the same settlement phase. In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal. Multiple phases at a single site were treated as independent of each other (Figure x).

As the excavation of archaeological sites typically follow from residential and commercial development and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances. In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like, using polygon features. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the rgrass7 package. In addition to code and original spatial data being available in the research compendia for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.

Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulation was performed. Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase. This year then has a corresponding likely elevation range of the contemporaneous shoreline. An elevation value was drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDist package for R. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.

horisontal and topographic distance between site and shore-line was measured.\
This process was repeated 1000 times for each phase for each site. The choice of 1000 iterations follows an evaluation of convergence on median distances using the example site Hovland 3. At 1000 simulation runs the analysis of each site ranged from hours, depending on the distance to the simulated shorelines and consequently the necessary size of the window of analysis.

The distances were then measured.

The occupation of a site and shifting sea-levels. The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials. Additionally, defining what shore bound settlement represents can be problematic. In the literature this is often qualified with statements along the lines of "the sites were located on or close to the contemporenous shore-line". What constitutes close is of course highly variable, complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line. Other elements of interest is the tidal range, and the exposure of the location in question, which can also have an impact on the distance of the site to the mean ocean surface.

The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

```{r get-data, eval = FALSE}
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

```{r demo-inline-code}
x <- round(pi, 2)
```

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
